Local PCA Shows How Population Structure Differs Along the Genome

### 2016-08-10 writeup/local_pca_paper.tex
added new figs; modified method figure

Dimension reduction techniques such as principal component analysis (PCA)
are often used to discover and visualize the patterns 
of mean kinship between individuals in large genomic datasets,
known as population structure.
% Modeling such kinship is key to controling the confounding effects of population structure in genome-wide association studies.
Genome-wide mean kinship
is an average of the relationships across all locus-specific genealogical trees,
which can be strongly affected on intermediate genomic scales by many biological factors,
such as linked selection.
We show how to use local principal components analysis to describe this meso-scale variation,
and apply the method to genomic data from three species,
finding in each that population structure varies on the scale of megabases to tens of megabases.
In a global human dataset, small, discontinuous variation is likely explained by polymorphic chromosomal inversions.
In a dataset of primarily African \textit{Drosophila melanogaster}, large, continuous variation across each chromosome arm
is explained by known chromosomal inversions thought to be under recent selection.
In a range-wide dataset of \textit{Medicago truncatula},
common axes of variation in population structure are shared between chromosomes,
correlate with local gene density,
and may be caused by background selection or local adaptation.
The method is implemented as an R package, called lostruct.

\plr{Add percent of variance explained by first two MDS coordinates, after subtracting the genome-wide mean covariance matrix.}

\plr{Add references to supplemental whole-genome PCA plots in the text.}

@@

Realized kinship summarizes the shapes
of the genealogical trees % that relate the samples
at each location along the genome.
Since these trees vary along the genome, so does realized kinship;
but averaging over sufficiently many trees we hope to get a stable estimate,
independent of the genomic region chosen.
This hope is not entirely justified: for instance,
kinship on sex chromosomes is expected to differ from the autosomes;
and positive or negative selection on particular loci can dramatically disort shapes of nearby genealogies
\citep{maynardsmith1974hitchhiking,barton2000genetic,neher2013genetic,charlesworth2012effects}.
Indeed,
many species show chromosome-scale variation in diversity and divergence
\citep[sometimes dramatic:][]{langley2012genomic};
species phylogenies can differ along the genome 
due to incomplete lineage sorting \citep[e.g.,][]{pease2013accurate},
adaptive introgression and/or local adaptation \citep[e.g.,][]{ellegren2012genomic,nadeau2012genomic,pool2015natural,vernot2014resurrecting};
and theoretical expectations predict that geographic patterns of relatedness should depend on selection
\citep{charlesworth2003review}.
Nonetheless, it is not generally known to what extent patterns of kinship vary along the genome,
nor for what reasons.

@@

@@

The general steps in our method could be carried out in many ways.
As depicted in Figure \ref{fig:diagram}, these are
(1) divide the genome into windows,
(2) summarize the patterns of relatedness (population structure) in each window,
(3) measure dissimilarity in population structure between each pair of windows,
(4) visualize the resulting dissimilarity matrix,
and (5) combine similar windows to more accurately estimate local population structure.
Details of how we carried these out are given below.

>>>>

Dimension reduction techniques such as principal component analysis (PCA)
are often used to discover and visualize the patterns 
of mean kinship between individuals in large genomic datasets,
\plr{known as population structure.}
% Modeling such kinship is key to controling the confounding effects of population structure in genome-wide association studies.
Mean kinship
is an average of the relationships across locus-specific genealogical trees,
which can be strongly affected on intermediate genomic scales by many biological factors,
such as linked selection.
We show how to use local principal components analysis to describe this meso-scale variation,
and apply the method to genomic data from three species,
finding in each that population structure varies on the scale of megabases to tens of megabases.
In a global human dataset, small, discontinuous variation is likely explained by polymorphic chromosomal inversions.
In a dataset of primarily African \textit{Drosophila melanogaster}, large, continuous variation across each chromosome arm
is explained by known chromosomal inversions thought to be under recent selection.
In a range-wide dataset of \textit{Medicago truncatula},
common axes of variation in population structure are shared between chromosomes,
correlate with local gene density,
and may be caused by background selection or local adaptation.
The method is implemented as an R package, called lostruct.
\plr{conclusion sentence.}
\plr{methods or results?}

\plr{check refs to supp figs}

@@

Realized kinship summarizes the shapes
of the genealogical trees % that relate the samples
at each location along the genome.
Since these trees vary along the genome, so does realized kinship;
but averaging over sufficiently many trees we hope to get a stable estimate,
that doesn't depend much on the genetic markers chosen.
This is not guarenteed:
for instance,
kinship on sex chromosomes is expected to differ from the autosomes;
and positive or negative selection on particular loci can dramatically disort shapes of nearby genealogies
\citep{maynardsmith1974hitchhiking,barton2000genetic,neher2013genetic,charlesworth2012effects}.
Indeed,
many species show chromosome-scale variation in diversity and divergence
\citep[sometimes dramatic:][]{langley2012genomic};
species phylogenies can differ along the genome 
due to incomplete lineage sorting \citep[e.g.,][]{pease2013accurate},
adaptive introgression and/or local adaptation \citep[e.g.,][]{ellegren2012genomic,nadeau2012genomic,pool2015natural,vernot2014resurrecting};
and theoretical expectations predict that geographic patterns of relatedness should depend on selection
\citep{charlesworth2003review}.
Nonetheless, it is not generally known to what extent patterns of kinship vary along the genome,
nor for what reasons.

@@

\plr{short paragraph describing what the approach does/brings to the table}

@@

As depicted in Figure \ref{fig:diagram}, the general steps to the method are:
(1) divide the genome into windows,
(2) summarize the patterns of relatedness (population structure) in each window,
(3) measure dissimilarity in population structure between each pair of windows,
(4) visualize the resulting dissimilarity matrix,
and (5) combine similar windows to more accurately estimate local population structure.
Details of how we carried these out are given below.
